---
title: "Biostat 203B Homework 3"
subtitle: "Due Feb 22 @ 11:59PM"
author: "Chih Wei Hsu, UID:706788612"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

Display machine information for reproducibility:
```{r}
sessionInfo()
```

Load necessary libraries (you can add more as needed).
```{r setup}
library(arrow)
library(gtsummary)
library(memuse)
library(pryr)
library(R.utils)
library(tidyverse)
library(lubridate)
```

Display your machine memory.
```{r}
memuse::Sys.meminfo()
```

In this exercise, we use tidyverse (ggplot2, dplyr, etc) to explore the [MIMIC-IV](https://physionet.org/content/mimiciv/3.1/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2026winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

## Q1. Visualizing patient trajectory

Visualizing a patient's encounters in a health care system is a common task in clinical data analysis. In this question, we will visualize a patient's ADT (admission-discharge-transfer) history and ICU vitals in the MIMIC-IV data.

### Q1.1 ADT history

A patient's ADT history records the time of admission, discharge, and transfer in the hospital. This figure shows the ADT history of the patient with `subject_id` 10001217 in the MIMIC-IV data. The x-axis is the calendar time, and the y-axis is the type of event (ADT, lab, procedure). The color of the line segment represents the care unit. The size of the line segment represents whether the care unit is an ICU/CCU. The crosses represent lab events, and the shape of the dots represents the type of procedure. The title of the figure shows the patient's demographic information and the subtitle shows top 3 diagnoses.

Do a similar visualization for the patient with `subject_id` 10063848 using ggplot.

Hint: We need to pull information from data files `patients.csv.gz`, `admissions.csv.gz`, `transfers.csv.gz`, `labevents.csv.gz`, `procedures_icd.csv.gz`, `diagnoses_icd.csv.gz`, `d_icd_procedures.csv.gz`, and `d_icd_diagnoses.csv.gz`. For the big file `labevents.csv.gz`, use the Parquet format you generated in Homework 2. For reproducibility, make the Parquet folder `labevents_pq` available at the current working directory `hw3`, for example, by a symbolic link. Make your code reproducible.

```{bash}
zcat < ~/mimic/hosp/patients.csv.gz | head -5
```
```{r}
subj_id <- 10063848
```
Load Data
```{r}
path <- "~/mimic/hosp"

patients <- read_csv(file.path(path, "patients.csv.gz"), 
  show_col_types = FALSE) |>
  filter(subject_id == subj_id)
admissions <- read_csv(file.path(path, "admissions.csv.gz"), 
  show_col_types = FALSE) |> 
  filter(subject_id == subj_id) |>
  print(width = Inf)
transfers <- read_csv(file.path(path, "transfers.csv.gz"), 
  show_col_types = FALSE) |> 
  filter(subject_id == subj_id)
diag <- read_csv(file.path(path, "diagnoses_icd.csv.gz"), 
  show_col_types = FALSE) |>
  filter(subject_id == subj_id)
d_diag <- read_csv(file.path(path, "d_icd_diagnoses.csv.gz"),
  show_col_types = FALSE)
proc <- read_csv(file.path(path, "procedures_icd.csv.gz"),
  show_col_types = FALSE) |>
  filter(subject_id == subj_id)
d_proc <- read_csv(file.path(path, "d_icd_procedures.csv.gz"),
  show_col_types = FALSE)

# Load labevents from Parquet 
labs <- open_dataset("../HW2/labevents_parquet", format = "parquet") |>
  select(subject_id, charttime) |>
  filter(subject_id == subj_id) |>
  collect()
```

```{r}
# Get unique race from admissions
patient_race <- admissions |>
  distinct(race) |>
  pull(race) |>
  first() 

# Title Info
patient_info <- patients |> 
  mutate(desc = paste0(
    "Patient ", subject_id, ", ", gender, ", ", 
    anchor_age, " years old", ", ", tolower(patient_race))) |>
  pull(desc)

# Top 3 Diagnoses
top_3_diag <- diag |>
  slice_head(n=3) |>  
  left_join(d_diag, by = c("icd_code", "icd_version")) |>
  pull(long_title) |>
  paste(collapse = "\n")

y_levels <- c("Procedure", "Lab", "ADT")

# ADT
adt_plot <- transfers |>
  filter(!is.na(careunit)) |>
  mutate(
    line_weight = if_else(str_detect(careunit, "ICU|CCU"), 5, 1.5),
    y_label = factor("ADT", levels = y_levels)
  )
# Lab
lab_pot <- labs |>
  mutate(y_label = factor("Lab", levels = y_levels))

# Procedure 
proc_plot <- proc |>
  left_join(d_proc, by = c("icd_code", "icd_version")) |>
  mutate(
    y_label = factor("Procedure", levels = y_levels),
    long_title_wrapped = str_wrap(long_title, width = 30)
    )
  
```


Visualization
```{r}
ggplot() +
  # ADT
  geom_segment(data = adt_plot, 
               aes(x = intime, xend = outtime, y = y_label, 
                   color = careunit, size = line_weight)) +
  # Procedures
  geom_point(data = proc_plot, 
    aes(x = chartdate, y = y_label, shape = long_title), size = 3) + 
  # Labs (Crosses)
  geom_point(data = labs, 
    aes(x = charttime, y = "Lab"), shape = 3) +
  # Aesthetics
  scale_size_identity() +
  labs(
    title = patient_info,
    subtitle = top_3_diag,
    x = "Calendar Time",
    y = NULL,
    color = "Care Unit",
    shape = "Procedure"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom", 
    legend.box = "vertical",
    legend.text = element_text(size = 8),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", size = 10)
  )+
  scale_y_discrete(drop = FALSE) +
  guides(shape = guide_legend(ncol = 2, byrow = TRUE))
```
### Q1.2 ICU stays

ICU stays are a subset of ADT history. This figure shows the vitals of the patient `10001217` during ICU stays. The x-axis is the calendar time, and the y-axis is the value of the vital. The color of the line represents the type of vital. The facet grid shows the abbreviation of the vital and the stay ID.


Do a similar visualization for the patient `10063848`.

```{r}
data_path <- "~/mimic/icu"
vitals_map <- tibble(
  itemid = c(220045, 220180, 220179, 220210, 223761),
  vital_name = c("HR", "NBPd", "NBPs", "RR", "Temperature F")
)
```

```{bash}
zcat < ~/mimic/icu/chartevents.csv.gz | head -4
```
```{bash}
zcat < ~/mimic/icu/icustays.csv.gz | head -4
```
Load and Filter Data
```{r}
# Load ICU stays to get intime/outtime windows
icustays <- read_csv(
  file.path(data_path, "icustays.csv.gz"), show_col_types = FALSE) |>
  filter(subject_id == subj_id) |>
  mutate(
    intime = as.POSIXct(intime) |> with_tz("UTC"),
    outtime = as.POSIXct(outtime) |> with_tz("UTC")
  ) |>
  select(subject_id, stay_id, intime, outtime)

vitals <- open_dataset(
  "../HW2/chartevents_parquet", format = "parquet") |>
  filter(subject_id == subj_id, itemid %in% vitals_map$itemid) |>
  collect() |>
  select(-any_of("stay_id")) |>
  mutate(
    charttime = as.POSIXct(charttime) |> 
      with_tz(tzone = "UTC")
  ) |>
  inner_join(icustays, by = "subject_id", relationship = "many-to-many") |>
  filter(charttime >= intime, charttime <= outtime) |>
  left_join(vitals_map, by = "itemid") |>
  mutate(vital_name = factor(
    vital_name, levels = vitals_map$vital_name)
    )
```
Visualization
```{r}
ggplot(vitals, aes(x = charttime, y = valuenum, color = vital_name)) +
  geom_line() +
  geom_point(size = 1) +
  facet_grid(vital_name ~ stay_id, scales = "free") +
  scale_color_brewer(palette = "Set1") +
  scale_x_datetime(
    guide = guide_axis(n.dodge = 2)) +
  labs(
    title = paste("Patient", subj_id, "ICU stays - Vitals"),
    x = "Calendar Time",
    y = "Value"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    strip.text.y = element_text(angle = -90),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    strip.background = element_rect(fill = "lightgrey"),
    strip.text.x = element_text(size = 10, face = "bold"),
    panel.grid.minor = element_blank()
  )
```
## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are
```{bash}
zcat < ~/mimic/icu/icustays.csv.gz | head -10
```

### Q2.1 Ingestion

Import `icustays.csv.gz` as a tibble `icustays_tble`.
```{r}
icustays_tble <- as_tibble(
  read_csv(file.path(data_path, "icustays.csv.gz"))
)
icustays_tble
```

### Q2.2 Summary and visualization

How many unique `subject_id`? 
```{r}
unique_subjects<- distinct(icustays_tble, subject_id) |>
  count()
unique_subjects
```
Can a `subject_id` have multiple ICU stays? 
```{r}
stays_summary <- icustays_tble |> 
  group_by(subject_id) |> 
  summarize(n_stays = n()) |> 
  ungroup()

any(stays_summary$n_stays > 1)
```
Summarize the number of ICU stays per `subject_id` by graphs. 
```{r}
ggplot(stays_summary, aes(x = n_stays)) +
  geom_bar(fill = "steelblue", color = "white") +
  scale_x_continuous(breaks = seq(1, max(stays_summary$n_stays), by = 1)) +
  labs(
    title = "Distribution of ICU Stays per Patient",
    subtitle = paste("Total Unique Patients:", unique_subjects),
    x = "Number of ICU Stays",
    y = "Number of Patients"
  ) +
  theme_minimal() 
```

## Q3. `admissions` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/hosp/admissions/> for details of each field in this file. The first 10 lines are
```{bash}
zcat < ~/mimic/hosp/admissions.csv.gz | head -10
```

### Q3.1 Ingestion

Import `admissions.csv.gz` as a tibble `admissions_tble`.

```{r}
admissions_tble <- as_tibble(
  read_csv('~/mimic/hosp/admissions.csv.gz')
)
admissions_tble

```

### Q3.2 Summary and visualization

Summarize the following information by graphics and explain any patterns you see.

- number of admissions per patient
```{r}
adm_per_patient <- admissions_tble |>
  count(subject_id, name = "n_admissions")

ggplot(adm_per_patient, aes(x = n_admissions)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    title = "Number of Admissions per Patient",
    x = "Number of Admissions",
    y = "Number of Patients"
  ) +
  theme_bw()
```
**solutions:**
The distribution is right-skewed. Most patients are admitted once, while a small proportion have multiple readmissions, suggesting a subset of chronically ill or high-risk patients.


- admission hour (anything unusual?)  
```{r}
admissions_htble <- admissions_tble |>
  mutate(admit_hour = hour(admittime))

ggplot(admissions_htble, aes(x = admit_hour)) +
  geom_histogram(binwidth = 1, fill = "darkorange", color = "white") +
  labs(
    title = "Admission Hour Distribution",
    x = "Hour of Day",
    y = "Number of Admissions"
  ) +
  theme_bw()

```
**solutions:**
Admissions tend to occur more frequently during daytime hours, possibly reflecting scheduled procedures and clinic referrals. A smaller peak around midnight may reflect administrative timestamping practices.

- admission minute (anything unusual?)  
```{r}
admissions_mtble <- admissions_tble |>
  mutate(admit_minute = minute(admittime))

ggplot(admissions_mtble, aes(x = admit_minute)) +
  geom_histogram(binwidth = 1, fill = "purple", color = "white") +
  labs(
    title = "Admission Minute Distribution",
    x = "Minute",
    y = "Number of Admissions"
  ) +
  theme_bw()
```
**solutions:**
A noticeable spike at minute 00, 15, 30, 45 suggests rounding or default timestamp recording. This indicates that admission times may not always reflect exact real-world times, but rather administrative or system-recorded times.

- length of hospital stay (from admission to discharge) (anything unusual?)  
```{r}
length_hstay <- admissions_tble |>
  mutate(
    los_days = as.numeric(dischtime - admittime, units = "days")
  )

ggplot(length_hstay, aes(x = los_days)) +
  geom_histogram(binwidth = 1, fill = "forestgreen", color = "white") +
  coord_cartesian(xlim = c(0, 30)) +  # zoom to see pattern
  labs(
    title = "Length of Hospital Stay (Days)",
    x = "Length of Stay (Days)",
    y = "Number of Admissions"
  ) +
  theme_bw()

```
**solutions:**
The length of stay distribution is heavily right-skewed. Most hospital stays are short (less than one week), while a small number of patients experience prolonged hospitalizations, possibly due to severe illness or complications.

According to the [MIMIC-IV documentation](https://mimic.mit.edu/docs/iv/about/concepts/#date-shifting), 

> All dates in the database have been shifted to protect patient confidentiality. Dates will be internally consistent for the same patient, but randomly distributed in the future. Dates of birth which occur in the present time are not true dates of birth. Furthermore, dates of birth which occur before the year 1900 occur if the patient is older than 89. In these cases, the patient’s age at their first admission has been fixed to 300.

## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/hosp/patients/> for details of each field in this file. The first 10 lines are
```{bash}
zcat < ~/mimic/hosp/patients.csv.gz | head -10
```

### Q4.1 Ingestion

Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/patients/>) as a tibble `patients_tble`.

```{r}
patients_tble <- as_tibble(
  read_csv('~/mimic/hosp/patients.csv.gz')
)
patients_tble

```

### Q4.2 Summary and visualization

Summarize variables `gender` and `anchor_age` by graphics, and explain any patterns you see.
```{r}
ggplot(patients_tble, aes(x = gender)) +
  geom_bar(fill = "steelblue", color = "white") +
  labs(
    title = "Gender Distribution",
    x = "Gender",
    y = "Number of Patients"
  ) +
  theme_bw()
```
**solutions:**
The gender distribution is relatively balanced, though slightly more female patients are observed. This reflects the general ICU population in MIMIC-IV, where female patients are somewhat more prevalent.

```{r}
ggplot(patients_tble, aes(x = anchor_age)) +
  geom_histogram(binwidth = 1, fill = "coral", color = "white") +
  labs(
    title = "Age Distribution at Admission",
    x = "Anchor Age",
    y = "Number of Patients"
  ) +
  theme_bw()
```
**solutions:**
The age distribution is concentrated among middle-aged and older adults, particularly between 50 and 70 years. The distribution is slightly right-skewed, with fewer very elderly patients. In MIMIC-IV, patients aged 90 and above are top-coded to 91 for privacy protection, which may create a subtle accumulation at age 91.

```{r}
ggplot(patients_tble, aes(x = anchor_age, fill = gender)) +
  geom_histogram(binwidth = 1, alpha = 0.5, position = "identity") +
  labs(
    title = "Anchor Age by Gender",
    x = "Anchor Age",
    y = "Count"
  ) +
  theme_bw()
```


## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are
```{bash}
zcat < ~/mimic/hosp/labevents.csv.gz | head -10
```

`d_labitems.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/d_labitems/>) is the dictionary of lab measurements. 
```{bash}
zcat < ~/mimic/hosp/d_labitems.csv.gz | head -10
```

We are interested in the lab measurements of albumin (50862), creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), and glucose (50931). Retrieve a subset of `labevents.csv.gz` that only containing these items for the patients in `icustays_tble`. Further restrict to the last available measurement (by `storetime`) before the ICU stay. The final `labevents_tble` should have one row per ICU stay and columns for each lab measurement.

Hint: Use the Parquet format you generated in Homework 2. For reproducibility, make `labevents_pq` folder available at the current working directory `hw3`, for example, by a symbolic link.

```{r}
library(arrow)
library(dplyr)
library(tidyr)


# 1) target itemids
lab_items <- c(
  50862, 50912, 50971, 50983, 50902, 50882, 51221, 51301, 50931
  )

lab_name <- c(
  albumin     = "50862",
  creatinine  = "50912",
  potassium   = "50971",
  sodium      = "50983",
  chloride    = "50902",
  bicarbonate = "50882",
  hematocrit  = "51221",
  wbc         = "51301",
  glucose     = "50931"
)

# 2) ICU reference table (keep only needed keys)
# IMPORTANT: make join keys consistent types (avoid int64 vs double)
icu_rtbl <- icustays_tble |>
  transmute(
    subject_id = as.integer(subject_id),
    hadm_id    = as.integer(hadm_id),
    stay_id,
    intime
  ) |>
  arrow_table()

# 3) Arrow pipeline: filter labs + join ICU + keep pre-ICU + last storetime
labs_last <- open_dataset("labevents_pq", format = "parquet") |>
  transmute(
    subject_id = as.integer(subject_id),
    hadm_id    = as.integer(hadm_id),
    itemid     = as.integer(itemid),
    storetime,
    valuenum
  ) |>
  filter(itemid %in% lab_items) |>
  inner_join(icu_rtbl, by = c("subject_id", "hadm_id")) |>
    mutate(
    storetime = arrow::cast(storetime, arrow::timestamp(unit = "us", timezone = "UTC")),
    intime    = arrow::cast(intime,    arrow::timestamp(unit = "us", timezone = "UTC"))
  ) |>
  filter(storetime <= intime) |>
  arrange(stay_id, itemid, desc(storetime)) |>
  distinct(stay_id, itemid, .keep_all = TRUE) |>
  collect()

id_to_name <- setNames(names(lab_name), lab_name)  # "50882" -> "bicarbonate"

labevents_tble <- labs_last |>
  transmute(stay_id, itemid = as.character(itemid), valuenum) |>
  pivot_wider(names_from = itemid, values_from = valuenum) |>
  arrange(stay_id)

id_cols <- intersect(names(id_to_name), names(labevents_tble))

labevents_tble <- labevents_tble |>
  rename_with(~ id_to_name[.x], .cols = all_of(id_cols))

labevents_tble
```

## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{bash}
zcat < ~/mimic/icu/chartevents.csv.gz | head -10
```

`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`. 
```{bash}
zcat < ~/mimic/icu/d_items.csv.gz | head -10
```

We are interested in the vitals for ICU patients: heart rate (220045), systolic non-invasive blood pressure (220179), diastolic non-invasive blood pressure (220180), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble`. Further restrict to the first vital measurement (by `storetime`) within the ICU stay. The final `chartevents_tble` should have one row per ICU stay and columns for each vital measurement. 


Hint: Use the Parquet format you generated in Homework 2. For reproducibility, make `chartevents_pq` folder available at the current working directory, for example, by a symbolic link.

**Solution**
```{r}
dchartitems_tble <- 
  read_csv('~/mimic/icu/d_items.csv.gz', show_col_types = FALSE) |>
  print(width = Inf)
```

```{r}
chartitem_itemid <- dchartitems_tble |>
  filter(itemid %in% c(
    220045, 
    220179, 
    220180, 
    223761, 
    220210
    )) |>
  print(width = Inf)
```
```{r}
name_map <- setNames(
  as.character(chartitem_itemid$itemid),  # 這是舊欄名
  str_to_lower(str_replace_all(chartitem_itemid$label, " ", "_"))  # 新欄名
)

t <- system.time({
  chartevents_tble <- open_dataset("chartevents_pq") |>
    to_duckdb() |>
    select(subject_id, stay_id, itemid, charttime, storetime, valuenum) |>
    filter(itemid %in% local(chartitem_itemid$itemid)) |>
    left_join(select(icustays_tble, stay_id, intime, outtime),
              by = "stay_id", copy = TRUE) |>
    filter(storetime >= intime, storetime <= outtime) |>
    group_by(subject_id, stay_id, itemid) |>
    slice_min(storetime, n = 1, with_ties = TRUE, na_rm = TRUE) |>
    summarise(valuenum = mean(valuenum, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(names_from = itemid, values_from = valuenum) |>
    collect()

chartevents_tble <- chartevents_tble |>
  rename_with(
    ~ str_to_lower(str_replace_all(
        chartitem_itemid$label[
          match(.x, chartitem_itemid$itemid)
        ],
        " ", "_"
      )),
    .cols = intersect(names(chartevents_tble),
                      as.character(chartitem_itemid$itemid))
  )
})

print(chartevents_tble, width = Inf)
```

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are all ICU stays of adults (age at `intime` >= 18) and columns contain at least following variables

- all variables in `icustays_tble`  
- all variables in `admissions_tble`  
- all variables in `patients_tble`
- the last lab measurements before the ICU stay in `labevents_tble` 
- the first vital measurements during the ICU stay in `chartevents_tble`

The final `mimic_icu_cohort` should have one row per ICU stay and columns for each variable.
```{r}
icu_adult <- icustays_tble |>
  left_join(
    select(patients_tble, subject_id, anchor_age, anchor_year),
    by = "subject_id"
  ) |>
  mutate(
    age_intime = anchor_age + (year(intime) - anchor_year)
  ) |>
  filter(age_intime >= 18)
```

```{r}
mimic_core <- icu_adult |>
  left_join(
    admissions_tble,
    by = "hadm_id",
    suffix = c("", "_adm")
  ) |>
  left_join(
    patients_tble,
    by = "subject_id",
    suffix = c("", "_pt")
  )
```

```{r}
mimic_icu_cohort <- mimic_core |>
  left_join(labevents_tble,   by = "stay_id") |>
  left_join(chartevents_tble, by = "stay_id")

print(mimic_icu_cohort, width = Inf)
```

## Q8. Exploratory data analysis (EDA)

Summarize the following information about the ICU stay cohort `mimic_icu_cohort` using appropriate numerics or graphs:

- Length of ICU stay `los` vs demographic variables (race, insurance, marital_status, gender, age at intime)

- Length of ICU stay `los` vs the last available lab measurements before ICU stay

- Length of ICU stay `los` vs the first vital measurements within the ICU stay

- Length of ICU stay `los` vs first ICU unit

## Q9. AI assistant

Which AI assistants (e.g., GitHub Copilot) do you use when working on this assignment? Which AI model (e.g., GPT-5 mini, GPT-5, Claude Sonnet 4.5) does the AI assistant use? How do you use them? Do you think they help improve your productivity? 

Give 5 instances where AI model gave incorrect or misleading answers. You can use screenshots or copy-paste the Q&A.

1. In Q1.2,  Error in `combine_vars()`:
! At least one layer must contain all faceting variables: `stay_id`
✖ Plot is missing `stay_id`
