---
title: "final_project_sol"
format: html
editor: visual
---

Display machine information:

```{r}
sessionInfo()
```

Display my machine memory.

```{r}
memuse::Sys.meminfo()
```

Load database libraries and the tidyverse frontend:

```{r}
library(bigrquery)
library(dbplyr)
library(DBI)
library(gt)
library(gtsummary)
library(tidyverse)
library(arrow)
library(data.table)
library(duckdb)
library(memuse)
library(pryr)
library(R.utils)
library(dplyr)
library(tidymodels)
library(ranger)
library(GGally)
```

```{r}
# path to the service account token 
satoken <- "biostat-203b-2026-winter-92fefbfab477.json"
# BigQuery authentication using service account
bq_auth(path = satoken)
```

```{r}
# connect to the BigQuery database `biostat-203b-2025-mimiciv_3_1`
con_bq <- dbConnect(
    bigrquery::bigquery(),
    project = "biostat-203b-2025-winter",
    dataset = "mimiciv_3_1",
    billing = "biostat-203b-2025-winter"
)
con_bq
```

List all tables in the `mimiciv_3_1` database.

```{r}
dbListTables(con_bq)
```

## corhort

```{r}
proc_tbl <- tbl(con_bq, "procedureevents") |>
  select(subject_id, hadm_id, stay_id, itemid, value, starttime, endtime) |>
  print()
```
```{r}
patients_tbl <- tbl(con_bq, "patients") |>
  select(subject_id, gender, anchor_age) |>
  arrange(subject_id) 
```

```{r}
cohort_proc <- proc_tbl |>
  filter(itemid == 225792) |>
  filter(!is.na(starttime), !is.na(endtime)) |>
  inner_join(patients_tbl |> 
  select(subject_id, anchor_age), by = "subject_id") |>
  filter(anchor_age > 65)

stay_map <- cohort_proc |>
  group_by(stay_id) |>
  summarise(
    subject_id = min(subject_id), 
    hadm_id = min(hadm_id), 
    .groups = "drop"
    )
```

divide group
```{r}
x1 <- cohort_proc |>
  mutate(start_dt = starttime, end_dt = endtime) |>
  filter(end_dt >= start_dt) |>
  mutate(
    run_end = sql("
      MAX(end_dt) OVER (
        PARTITION BY stay_id
        ORDER BY start_dt, end_dt
        ROWS BETWEEN UNBOUNDED PRECEDING AND CURRENT ROW
      )
    ")
  )

x2 <- x1 |>
  mutate(
    prev_run_end = sql("
      LAG(run_end) OVER (
        PARTITION BY stay_id
        ORDER BY start_dt, end_dt
      )
    ")
  )

x3 <- x2 |>
  mutate(
    grp_id = sql("
      SUM(IF(prev_run_end IS NULL OR start_dt > prev_run_end, 1, 0))
      OVER (PARTITION BY stay_id ORDER BY start_dt, end_dt)
    ")
  )
```

```{r}
cut_hr <- 48

cohort_tbl <- x3 |>
  group_by(stay_id, grp_id) |>
  summarise(
    grp_start = min(start_dt),
    grp_end   = max(end_dt),
    .groups = "drop"
  ) |>
  mutate(grp_hr = sql("DATETIME_DIFF(grp_end, grp_start, SECOND)") / 3600.0) |>
  group_by(stay_id) |>
  summarise(
    vent_start = min(grp_start),
    vent_end   = max(grp_end),
    vent_hr    = sum(grp_hr),
    n_vent_segments = n(),
    y_prolong  = sql(paste0("CAST(SUM(grp_hr) > ", cut_hr, " AS INT64)")),
    .groups = "drop"
  ) |>
  left_join(stay_map, by = "stay_id")
```

```{r}
cohort_tbl |>
  summarise(
    n = n(),
    n_stay = n_distinct(stay_id),
    n_subj = n_distinct(subject_id),
    prop_prolong = avg(y_prolong)
  ) |>
  collect()
```

```{r}
cohort_tbl |>
  head(20) |>
  collect()
```




## feature


### admissions

```{r}
admissions_tbl <- tbl(con_bq, "admissions") |>
  select(
    subject_id, 
    hadm_id, 
    admission_type, 
    admission_location, 
    race, 
    hospital_expire_flag
    ) |>
  arrange(subject_id, hadm_id) |>
  print(width=inf)
```

### icustays

```{r}
icustays_tbl   <- tbl(con_bq, "icustays") |>
  arrange(subject_id, stay_id) |>
  print(width=inf)
```

### omr


```{r}
omr_tbl <- tbl(con_bq, "omr") |>
  arrange(subject_id) |>
  print(width=inf)
```

```{r}
# helper: base OMR with stay anchor
omr_base <- cohort_tbl |>
  select(stay_id, subject_id, vent_start = vent_start) |>
  inner_join(
    omr_tbl |>
      filter(result_name %in% c("BMI (kg/m2)", "Blood Pressure")),
    by = "subject_id"
  ) |>
  mutate(
    # distance in seconds (DATETIME vs DATE)
    diff_sec = sql("ABS(DATETIME_DIFF(DATETIME(chartdate), DATETIME(vent_start), SECOND))")
  )

# 1) BMI nearest per stay
omr_bmi_near <- omr_base |>
  filter(result_name == "BMI (kg/m2)") |>
  mutate(
    rn = sql("
      ROW_NUMBER() OVER(
        PARTITION BY stay_id
        ORDER BY diff_sec ASC, chartdate DESC, seq_num DESC
      )
    ")
  ) |>
  filter(rn == 1) |>
  transmute(
    stay_id,
    bmi = sql("SAFE_CAST(result_value AS FLOAT64)")
  )

# 2) BP nearest per stay
omr_bp_near <- omr_base |>
  filter(result_name == "Blood Pressure") |>
  mutate(
    rn = sql("
      ROW_NUMBER() OVER(
        PARTITION BY stay_id
        ORDER BY diff_sec ASC, chartdate DESC, seq_num DESC
      )
    ")
  ) |>
  filter(rn == 1) |>
  transmute(
    stay_id,
    bp_raw = result_value,
    sbp = sql("SAFE_CAST(SPLIT(result_value,'/')[OFFSET(0)] AS FLOAT64)"),
    dbp = sql("SAFE_CAST(SPLIT(result_value,'/')[OFFSET(1)] AS FLOAT64)")
  )

# 3) final OMR stay-level features (1 row per stay_id)
omr_tbl <- cohort_tbl |>
  select(stay_id) |>
  left_join(omr_bmi_near, by = "stay_id") |>
  left_join(omr_bp_near,  by = "stay_id")
```

```{r}
omr_tbl |>
  summarise(
    n = n(),
    miss_bmi = sql("AVG(IF(bmi IS NULL, 1, 0))"),
    miss_sbp = sql("AVG(IF(sbp IS NULL, 1, 0))"),
    miss_dbp = sql("AVG(IF(dbp IS NULL, 1, 0))")
  ) |>
  collect()
```



### labevents

```{bash}
zcat < ~/mimic/hosp/d_labitems.csv.gz | grep -i Nitrogen
```

```{r}
lab_items <- c(
   51006, 50912, 50971, 50983
  )

labevents_tbl <- tbl(con_bq, "labevents") |>
  filter(itemid %in% lab_items) |>
  
  inner_join(
    cohort_proc |>
      select(subject_id, hadm_id, stay_id, vent_start = starttime),
    by = c("subject_id", "hadm_id")
  ) |>
  filter(!is.na(storetime), !is.na(valuenum)) |>
  

  filter(sql("DATETIME(storetime)") <= vent_start) |>
  group_by(stay_id, itemid) |>
  mutate(last_storetime = max(storetime)) |>
  ungroup() |>
  filter(storetime == last_storetime) |>

  # avg
  group_by(stay_id, itemid) |>
  summarise(valuenum = avg(valuenum), .groups = "drop") |>
  
  group_by(stay_id) |>
  summarise(
    UreaNitrogen = sql("MAX(IF(itemid=51006, valuenum, NULL))"),
    creatinine   = sql("MAX(IF(itemid=50912, valuenum, NULL))"),
    potassium    = sql("MAX(IF(itemid=50971, valuenum, NULL))"),
    sodium       = sql("MAX(IF(itemid=50983, valuenum, NULL))"),
    .groups = "drop"
  ) |>
  print()
```


### chartevents

```{r}
d_items_tbl <- tbl(con_bq, "d_items")
```

```{r}
chart_items <- c(
    220045,   # heart rate
    220179,   # systolic_bp
    220180,   # diastolic_bp
    223761,   # temperature_c
    220210    # respiratory_rate
  ) 

chartevents_tbl <- tbl(con_bq, "chartevents") |>
  # 1) restrict to cohort stays + bring in anchor time (starttime)
  inner_join(
    cohort_proc |>
      select(stay_id, starttime),  
    by = "stay_id"
  ) |>
  # 2) keep target vitals only
  filter(itemid %in% chart_items) |>
  filter(!is.na(storetime), !is.na(valuenum)) |>
  # 3) keep events at/after starttime
  #    storetime is usually TIMESTAMP; starttime may be DATETIME -> cast storetime to DATETIME
  filter(sql("DATETIME(storetime)") <= starttime) |>
  # 4) find first storetime per stay_id x itemid
  group_by(stay_id, itemid) |>
  mutate(first_storetime = min(storetime)) |>
  ungroup() |>
  filter(storetime == first_storetime) |>
  # 5) average ties at first storetime
  group_by(stay_id, itemid) |>
  summarise(
    valuenum = avg(valuenum),   # BigQuery AVG ignores NULL; don't use na.rm
    .groups = "drop"
  ) |>
  # 6) pivot to wide columns (one row per stay_id)
  group_by(stay_id) |>
  summarise(
    heart_rate       = sql("MAX(IF(itemid=220045, valuenum, NULL))"),
    systolic_bp      = sql("MAX(IF(itemid=220179, valuenum, NULL))"),
    diastolic_bp     = sql("MAX(IF(itemid=220180, valuenum, NULL))"),
    temperature_c    = sql("MAX(IF(itemid=223761, valuenum, NULL))"),
    respiratory_rate = sql("MAX(IF(itemid=220210, valuenum, NULL))"),
    .groups = "drop"
  ) |>
  print()
```

### services

```{r}
services_base <- tbl(con_bq, "services") |>
  select(subject_id, hadm_id, transfertime, curr_service)
```

```{r}
services_last_time <- services_base |>
  group_by(subject_id, hadm_id) |>
  summarise(
    last_transfertime = max(transfertime),
    .groups = "drop"
  )

services_tbl <- services_base |>
  inner_join(services_last_time, by = c("subject_id","hadm_id")) |>
  filter(transfertime == last_transfertime) |>
  group_by(subject_id, hadm_id) |>
  summarise(
    curr_service = sql("ANY_VALUE(curr_service)"),
    .groups = "drop"
  )
```

## putting things together

```{r}
model_tbl <- cohort_tbl |>
  left_join(admissions_tbl, by = c("subject_id","hadm_id")) |>
  left_join(patients_tbl,   by = "subject_id") |>
  left_join(icustays_tbl |> select(stay_id, intime, outtime, first_careunit, last_careunit),
            by = "stay_id") |>
  left_join(omr_tbl,   by = "stay_id") |>
  left_join(services_tbl,   by = c("subject_id","hadm_id")) |>
  left_join(labevents_tbl,  by = "stay_id") |>
  left_join(chartevents_tbl,by = "stay_id") |>
  arrange(subject_id) |>
  collect() |>
  print(width =inf) 
```




```{r}
check_keys <- function(x, name){
  x |>
    summarise(
      n = n(),
      n_subj = n_distinct(subject_id),
      n_hadm = n_distinct(hadm_id),
      n_stay = n_distinct(stay_id)
    ) |>
    collect() |>
    mutate(step = name)
}

check_keys(cohort_proc, "cohort_proc")      
check_keys(cohort_tbl, "cohort_tbl")
check_keys(model_tbl,   "after all joins")

```

```{r}
miss_tbl <- model_tbl |>
  summarise(
    n = n(),

    miss_urea = sql("AVG(IF(UreaNitrogen IS NULL, 1, 0))"),
    miss_cr   = sql("AVG(IF(creatinine IS NULL, 1, 0))"),
    miss_k    = sql("AVG(IF(potassium IS NULL, 1, 0))"),
    miss_na   = sql("AVG(IF(sodium IS NULL, 1, 0))"),

    miss_hr   = sql("AVG(IF(heart_rate IS NULL, 1, 0))"),
    miss_sbp  = sql("AVG(IF(systolic_bp IS NULL, 1, 0))"),
    miss_dbp  = sql("AVG(IF(diastolic_bp IS NULL, 1, 0))"),
    miss_temp = sql("AVG(IF(temperature_c IS NULL, 1, 0))"),
    miss_rr   = sql("AVG(IF(respiratory_rate IS NULL, 1, 0))")
  ) |>
  collect()
miss_tbl

```
```{r}
model_local <- model_tbl |>
  select(
    stay_id, subject_id, y_prolong,
    anchor_age, gender, race,
    hospital_expire_flag,
    UreaNitrogen, creatinine, potassium, sodium,
    heart_rate, systolic_bp, diastolic_bp, temperature_c, respiratory_rate,
    bmi, sbp, dbp
  ) |>
  collect()
```


```{r}
miss_rate <- sapply(model_local, function(x) mean(is.na(x)))
miss_rate <- sort(miss_rate, decreasing = TRUE)
miss_rate
```




## prep

```{r}
df <- model_tbl |>
  mutate(y_prolong = factor(y_prolong, levels = c(0, 1))) |>
  select(-any_of(c(
    "stay_id","subject_id","hadm_id",
    "intime","outtime",
    "vent_start","vent_end","vent_start_min","vent_end_max",
    "n_vent_events",
    "omr_date","bp_raw",
    "vent_hr","dur_hr","dur_sec"
  ))) |>
  mutate(across(where(is.character), ~ na_if(.x, ""))) |>
  select(y_prolong, where(is.numeric), where(is.factor), where(is.character))

# spilt
set.seed(203)
split_obj <- initial_split(df, prop = 0.8, strata = y_prolong)
train_df  <- training(split_obj)
test_df   <- testing(split_obj)

dim(train_df)
dim(test_df)

```

## Logit modeling

### Recipe: missing indicator + median impute
```{r}
logit_recipe <- recipe(y_prolong ~ ., data = train_df) |>
  step_zv(all_predictors()) |>
  step_string2factor(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors(), new_level = "Unknown") |>
  step_indicate_na(all_numeric_predictors()) |>
  step_impute_median(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>
  print()
```

### Model + workflow
```{r}
logit_mod <- 
  logistic_reg(
    penalty = tune(), 
    mixture = tune()
  ) |>
  set_engine("glmnet", standardize = FALSE) |>
  print()

logit_wf <- workflow() |>
  add_recipe(logit_recipe) |>
  add_model(logit_mod) |>
  print()

```

### Tuning grid
```{r}
logit_param_grid <- grid_regular(
  penalty(range = c(-6, 3)), 
  mixture(),
  levels = c(100, 5)
  ) |>
  print()
```


### CV 
```{r}
# Set cross-validation partitions.
set.seed(203)

logit_folds <- vfold_cv(train_df, v = 5, strata = y_prolong)
logit_folds

# Fit cross-validation.
(logit_fit <- logit_wf |>
  tune_grid(
    resamples = logit_folds,
    grid = logit_param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )) |>
  system.time()
```

```{r}
logit_fit
```

### Visualize CV results
```{r}
logit_fit |>
  # aggregate metrics from K folds
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = penalty, y = mean, color = factor(mixture))) +
  geom_point() +
  labs(x = "Penalty", y = "CV AUC") +
  scale_x_log10()
```

### Show the top 5 models

```{r}
logit_fit |>
  show_best(metric = "roc_auc")

```


### Select the best model.

```{r}
best_logit <- logit_fit |>
  select_best(metric = "roc_auc")
best_logit
```

### Logit Final workflow

```{r}
logit_final_wf <- logit_wf |>
  finalize_workflow(best_logit)
logit_final_wf
```


### Fit the whole training set, then predict the test cases

```{r}
logit_final_fit <- 
  logit_final_wf |>
  last_fit(split_obj)
logit_final_fit
```
### Test metrics
```{r}
logit_final_fit |> 
  collect_metrics()
```



## RF modeling

### Recipe

```{r}
rf_recipe <- recipe(y_prolong ~ ., data = train_df) |>
  step_zv(all_predictors()) |>
  step_string2factor(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors(), new_level = "Unknown") |>
  step_indicate_na(all_numeric_predictors()) |>
  step_impute_median(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE)
```


### model 

```{r}
rf_mod <- 
  rand_forest(
    mode = "classification",
    # Number of predictors randomly sampled in each split
    mtry = tune(),
    # Number of trees in ensemble
    trees = tune()
  ) |> 
  set_engine("ranger")
rf_mod
```

### Workflow

```{r}
rf_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_mod)
rf_wf
```

### tuning grid

```{r}
rf_param_grid <- grid_regular(
  trees(range = c(100L, 500L)), 
  mtry(range = c(1L, 5L)),
  levels = c(5, 5)
  )
rf_param_grid
```

### CV

```{r}
 # Set cross-validation partitions.
set.seed(203)

rf_folds <- vfold_cv(train_df, v = 5, strata = y_prolong)
rf_folds

rf_fit <- rf_wf |>
  tune_grid(
    resamples = rf_folds,
    grid = rf_param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
rf_fit
```

### Visualize CV results

```{r}
rf_fit |>
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = trees, y = mean, color = factor(mtry))) +
  geom_point() + 
  # geom_line() + 
  labs(x = "Num. of Trees", y = "CV AUC")
```


### Show the top 5 models.

```{r}
rf_fit |>
  show_best(metric = "roc_auc")
```

### select best model

```{r}
best_rf <- rf_fit |>
  select_best(metric = "roc_auc")
best_rf
```

### RF Final workflow

```{r}
rf_final_wf <- rf_wf |>
  finalize_workflow(best_rf)
rf_final_wf
```

### Fit the whole training set, then predict the test cases

```{r}
rf_final_fit <- 
  rf_final_wf |>
  last_fit(split_obj)
rf_final_fit
```

### Test metrics

```{r}
rf_final_fit |> 
  collect_metrics()
```

